Recent demography drives changes in linked selection across the maize genome

### 2015-11-10 manuscript.tex
edits

\begin{abstract}
Genetic drift and selection interact to determine the level of genetic variability present in a population.
Linked selection, whereby polymporhisms at neutral sites are manipulated by selection at linked sites, plays a primary roll in shaping genetic diversity. However, patterns of linked selection depends heavily on a population's propensity to drift, which in turn is shaped by its demographic history. Generally, demographic history is accounted for in terms of long-term effective population size ($N_e$). In practice, bottlenecks, invasions, or novel selective pressures can cause demography to change rapidly and dramatically, but the immediate genomic effects of a changing demography have not been well studied.   To address this, We investigated linked selection in in the genomes of 23 domesticated maize and 13 wild maize (teosinte) individuals. We show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid expansion post-domestication. We observe that hard sweeps on genic mutations are not the primary force driving maize evolution. As expected, a reduced population size during domestication decreased the efficiency of purifying selection to purge deleterious alleles from maize, but rapid expansion after domestication has increased the efficiency of purifying selection to levels exceeding those seen in teosinte. This final observation demonstrates that rapid demographic change can have wide-ranging impacts on diversity that conflict with would-be expectations based on long-term $N_e$.  
\end{abstract}

@@

The absence of evidence for a genome-wide impact of hard sweeps in coding regions differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}, but is consistent with estimates of the percentage of nonsynonmyous substitutions fixed by natural selection \cite{ross2009,sella2009,eyre2009estimating,williamson2014}. 
While larger long-term $N_e$ likely explains some of this difference, we see little change in the importance of hard sweeps in genes in singleton diversity in modern maize (Figure \ref{sFig:singleton}, suggesting other factors may contribute to these differences.  
Another possibility is that, if mutational target size scales with genome size, the larger genomes of human and maize may offer more opportunities for noncoding loci to contribute to adaptation, with hard sweeps on nonsynonymous variants then playing a relatively smaller role. 
Support from this idea comes from numerous cases of adaptive transposable element insertion modifying gene regulation \cite{studer2011,castelletti2014mite,mao2015,yang2013} and studies of local adaptation that show enrichment for SNPs in regulatory regions in teosinte \cite{pyhajarvi2013} and humans \cite{fraser2013gene} but for nonsynonymous variants in the small \emph{Arabidopsis} genome.

@@

>>>>

\begin{abstract}
Genetic drift and selection interact to determine the level of genetic variability present in a population.
Linked selection, whereby polymorphisms at neutral sites are manipulated by selection at linked sites, plays a primary roll in shaping genetic diversity. However, patterns of linked selection depend heavily on a population's propensity to drift, which in turn is shaped by its demographic history. Generally, demographic history is accounted for in terms of long-term effective population size ($N_e$). In practice, bottlenecks, invasions, or novel selective pressures can cause demography to change rapidly and dramatically, but the immediate genomic effects of a changing demography have not been well studied.   To address this, We investigated linked selection in in the genomes of 23 domesticated maize and 13 wild maize (teosinte) individuals. We show that maize went through a domestication bottleneck with a population size of approximately 5\% that of teosinte before it experienced rapid expansion post-domestication. We observe that hard sweeps on genic mutations are not the primary force driving maize evolution. As expected, a reduced population size during domestication decreased the efficiency of purifying selection to purge deleterious alleles from maize, but rapid expansion after domestication has increased the efficiency of purifying selection to levels exceeding those seen in teosinte. This final observation demonstrates that rapid demographic change can have wide-ranging impacts on diversity that conflict with would-be expectations based on long-term $N_e$.  
\end{abstract}

@@

The absence of evidence for a genome-wide impact of hard sweeps in coding regions differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}.
Comparisons of estimates of the percentage of nonsynonmyous substitutions fixed by natural selection \cite{ross2009,sella2009,eyre2009estimating,williamson2014} give similar results. 
While larger long-term $N_e$ likely explains some of this difference, we see little change in the importance of hard sweeps in genes in singleton diversity in modern maize (Figure \ref{sFig:singleton}, perhaps suggesting other factors may contribute to these differences.  
One possibility for example, is that if mutational target size scales with genome size, the larger genomes of human and maize may offer more opportunities for noncoding loci to contribute to adaptation, with hard sweeps on nonsynonymous variants then playing a relatively smaller role. 
Support from this idea comes from numerous cases of adaptive transposable element insertion modifying gene regulation in maize \cite{studer2011,castelletti2014mite,mao2015,yang2013} and studies of local adaptation that show enrichment for SNPs in regulatory regions in teosinte \cite{pyhajarvi2013} and humans \cite{fraser2013gene} but for nonsynonymous variants in the small \emph{Arabidopsis} genome.  
Our results, for example, are not very different from those of \cite{Halligan:2013ir}, who find no differences in diversity around nonsynonymous and synonymous substitutions and estimate that as many as 80\% of adaptive substitutions occur outside of genes.  
Future comparative analyses using a common statistical framework (e.g. \cite{corbett2015}) should allow explicit testing of this idea.

@@

\cite{Elyashiv:2010ga} $N_e$ among populations leads to differences in efficiency of purifying selection
